---
title: "Simulations In-Class Project"
date: "Due October 26, 2018 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:



With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 


Since you all have completed problem 1, you first step will be to work through each of your groupmates code for problem #1 and comment on what is happening. Then as a team move forward in on direction as you move on to the rest of the problems 2-5

1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    
```{r}
# play function plays a fair game with a maximum number of bets, a set bet value, a bankroll, and an output argument to specify the output of the function (default output is bankroll)

play <- function(maxNumberOfBets, betValue, bankroll, output = "bankroll") {
  betCount <- 0 # initialize bet counter
  result <- matrix(NA, nrow=1, ncol=2) # initialize matrix to store betCount and bankroll final variables in case the user would like both variables outputted (specifying anything other than "bankroll" or "bet count" as the argument)
  for (i in 1:maxNumberOfBets) { # for every bet up to the maximum number of bets
    if (bankroll >= betValue) { # if the player has enough money to make the bet, plau - if not, break.
      betCount <- betCount + 1 # the bet is placed, and the bet counter increases by one
      if (rbinom(1, 1, 0.5) == 1) { # the game is played, with a random result of success or failure - 1 indicates success (you win!)
        bankroll <- bankroll + betValue # if you win, the bet amount is added to your bankroll
      } else { 
        bankroll <- bankroll - betValue # if the result is not a 1, you lose, and the bet amount is subtracted from your bankroll
      }
    } else {
      break
    }
  }
  result[1,1] <- betCount # update the results matrix
  result[1,2] <- bankroll
  
  if (output=="bankroll"){ # output the value the user specified, or both if alternative key given
    return(result[1,2])
  } else if (output=="bet count") {
    return(result[1,1])
  } else {
    return(result)
  }
}

bustedProb <- (sum(replicate(5000, (play(100,100,1000)==0))) / 5000)
print(bustedProb)

```

b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    
```{r}
bustedProb2 <- (sum(replicate(5000, (play(500, 100, 1000)==0))) / 5000)
print(bustedProb2)

```

c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}

meanBust <- function(numberOfHands) {
  bustNums <- vector(length = 0)
  bustCount <- replicate(numberOfHands, play(100, 100, 1000, output="bet count"))
  for (i in 1:length(bustCount)) {
    if (bustCount[i] < 100) {
      bustNums <- c(bustNums, bustCount[i])
    }
  }
  return(mean(bustNums))
}

meanBust(5000)
```


d. the mean and variance of your bankroll after 100 hands (including busts).

```{r}

bankrollMeanVar <- function(numberOfHands) {
  bankrollTotals <- replicate(numberOfHands, play(100, 100, 1000))
  bankrollMean <- mean(bankrollTotals)
  bankrollVar <- var(bankrollTotals)
  print(paste("The bankroll mean after", numberOfHands, "hands is: $", bankrollMean))
  print(paste("The bankroll variance after", numberOfHands, "hands is: $", round(bankrollVar, 2)))
}

bankrollMeanVar(100)
```

e. the mean and variance of your bankroll after 500 hands (including busts).

```{r}
bankrollMeanVar(500)
```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.

```{r}
# play2 function plays a game where the probability of success is 18/38 with a maximum number of bets, a set bet value, a bankroll, and an output argument to specify the output of the function (default output is bankroll)

play2 <- function(maxNumberOfBets, betValue, bankroll, output = "bankroll") {
  betCount <- 0 # initialize bet counter
  result <- matrix(NA, nrow=1, ncol=2) # initialize matrix to store betCount and bankroll final variables in case the user would like both variables outputted (specifying anything other than "bankroll" or "bet count" as the argument)
  for (i in 1:maxNumberOfBets) { # for every bet up to the maximum number of bets
    if (bankroll >= betValue) { # if the player has enough money to make the bet, plau - if not, break.
      betCount <- betCount + 1 # the bet is placed, and the bet counter increases by one
      if (rbinom(1, 1, 18/38) == 1) { # the game is played, with a random result of success or failure - 1 indicates success (you win!)
        bankroll <- bankroll + betValue # if you win, the bet amount is added to your bankroll
      } else { 
        bankroll <- bankroll - betValue # if the result is not a 1, you lose, and the bet amount is subtracted from your bankroll
      }
    } else {
      break
    }
  }
  result[1,1] <- betCount # update the results matrix
  result[1,2] <- bankroll
  
  if (output=="bankroll"){ # output the value the user specified, or both if alternative key given
    return(result[1,2])
  } else if (output=="bet count") {
    return(result[1,1])
  } else {
    return(result)
  }
}

bustedProb <- (sum(replicate(5000, (play2(100,100,1000)==0))) / 5000)
print(bustedProb)

```

b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    
```{r}
bustedProb2 <- (sum(replicate(5000, (play(500, 100, 1000)==0))) / 5000)
print(bustedProb2)

```

c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
meanBust <- function(numberOfHands) {
  bustNums <- vector(length = 0)
  bustCount <- replicate(numberOfHands, play2(100, 100, 1000, output="bet count"))
  for (i in 1:length(bustCount)) {
    if (bustCount[i] < 100) {
      bustNums <- c(bustNums, bustCount[i])
    }
  }
  return(mean(bustNums))
}

meanBust(5000)
```


d. the mean and variance of your bankroll after 100 hands (including busts).

```{r}

bankrollMeanVar <- function(numberOfHands) {
  bankrollTotals <- replicate(numberOfHands, play2(100, 100, 1000))
  bankrollMean <- mean(bankrollTotals)
  bankrollVar <- var(bankrollTotals)
  print(paste("The bankroll mean after", numberOfHands, "hands is: $", bankrollMean))
  print(paste("The bankroll variance after", numberOfHands, "hands is: $", round(bankrollVar, 2)))
}

bankrollMeanVar(100)
```

e. the mean and variance of your bankroll after 500 hands (including busts).

```{r}
bankrollMeanVar(500)
```
 


3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    
```{r}

markovGame <- function(numberOfHands, betValue, bankroll, starting_p = 0.48, increment = 0.01) {
  # the number of hands, value of the bet, initial bankroll, starting probability of winning, and the increment are entered as the initial arguments
  
  p <- starting_p # the initial probability is the value provided in the arguments
  
  for (i in 1:numberOfHands) { # for every hand, up to the maximum number of hands
    if (rbinom(1, 1, p) == 1) { # play the hand, and if the result is a 1 (a win!)
      bankroll <- bankroll + betValue # add the bet value to the bankroll
      if (p < 1 - increment) {
        p <- (p + increment)
      } 
    } else { # if the result is not a 1 (you lose!), and the probability returns to the starting probability, and you lose your bet
      p <- starting_p
      bankroll <- bankroll - betValue
    } 
  }
  return(bankroll)
}

print(markovGame(100000, 100, 500000))

expect48 <- as.vector(replicate(100, markovGame(100000, 100, 500000)))
print(expect48)
mean(expect48)

# It is not a fair game because we would expect our value to be (within 2% or) very close to our initial bankroll.
```

b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?

```{r}

return100K50 <- markovGame(100000, 100, 5000000, starting_p = 0.48)
print(return100K50)

bestP100Hands <- function(numberOfHands, betValue, bankroll, starting_p = 0.48, increment = 0.01) {
  high <- starting_p+0.02
  low <- starting_p-0.02
  mid <- starting_p
  
  expectValue <- mean(as.vector(replicate(100, markovGame(numberOfHands, betValue, bankroll, mid))))

  while (!((expectValue <= bankroll*1.02) & (expectValue >= bankroll*0.98))) {
    if (expectValue > bankroll) {
      high <- mid
      mid <- (low+high)/2
    } else {
      low <- mid
      mid <- (low+high)/2
    }
      expectValue <- mean(as.vector(replicate(100, markovGame(numberOfHands, betValue, bankroll, mid))))
      print(paste("p is", mid))
      print(paste("expectedValue is", expectValue))
  }
  print(mid)
}

bestP100Hands(10000, 100, 500000, 0.48)

```

c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 

```{r}

expectChanged <- as.vector(replicate(100, markovGame(100000, 100, 500000, starting_p = 0.48, increment = 0.03)))
print(expectChanged)
mean(expectChanged)


bestIncr100Hands <- function(numberOfHands, betValue, bankroll, starting_p = 0.48, increment = 0.01) {
  expectValue <- mean(as.vector(replicate(100, markovGame(numberOfHands, betValue, bankroll, starting_p, increment))))

  while (!((expectValue <= bankroll*1.02) & (expectValue >= bankroll*0.98))) {
    if (increment < 1.00) {
      if (expectValue > bankroll*1.02) {
        increment <- increment - 0.01
      } else {
        increment <- increment + 0.01
      }
      expectValue <- mean(as.vector(replicate(100, markovGame(numberOfHands, betValue, bankroll, starting_p, increment))))
      print(paste("increment is", increment))
      print(paste("expectedValue is", expectValue))
    } else {
      break
    }
  }
  print(increment)
  print(expectValue)
}

bestIncr100Hands(1000, 100, 500000, 0.48, 0.11)
```


4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
```{r}
my_data <- sample(1:10, 10, replace = T)
my_data
my_sample <- my_data[sample(my_data, length(my_data), replace = T)]
my_sample
```
  
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
```{r}
sample_mean = mean(my_sample, na.rm = T)
sample_mean
sample_var = var(my_sample, na.rm = T)
sample_var
```
  
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
```{r}
a=replicate(1000,my_data[sample(my_data, length(my_data), replace = T)])# repeat 1000 times, each column means one sample


mean1000=colMeans(a) # mean for each column
mean1000

var1000=apply(a,2,var)# variance for each column
var1000
```
  
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile
```{r}
CI.LL <- qnorm(0.025,length(my_data) - 1)*sd(my_data)/sqrt(length(my_data))
CI.UL <- qnorm(0.975,length(my_data) - 1)*sd(my_data)/sqrt(length(my_data))
```

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 
```{r}

```


5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?